A template for the \emph{arxiv} style

### 2022-09-17 paper-rmd/paper-papaja-revision.Rmd
supplement revisions

keywords          : "probability judgments, Bayesian cognitive science, heuristics and biases"
wordcount         : "????"

@@

By assumption, a maximum of 50% of samples can be misread on average, so $d$ is a number in the range $[0, 1/2]$. The overall consequence of the sample-reading noise will be to shrink probability estimates toward .50 in proportion to $d$. The PT+N theory provides a unified account for a wide variety of biases in probability judgment that were previously attributed to different types of heuristics, as well as novel biases identified based on the model's predictions [@costello.watts2014; @costello.watts2016; @costello.watts2017; @costello.watts2018]. For example, the PT+N theory offers an explanation for many instances of "conservatism" [@costello.watts2014]---people's tendency to shy away from extreme probability judgments near 0 and 1, even when strong evidence warrants such judgments [e.g. @edwards.1968; @erev.etal1994].

@@

These models make subtly different predictions on two levels: First,  the models have distinct accounts of conditional probability judgments that make different predictions in terms of expected values. Second, the models present different process-level accounts of probability judgment that entail different predictions about the shape of the distribution of responses across individuals.

@@

In contrast, the PT+N model presents a constructive account of conditional probability judgments that is fundamentally non-Bayesian [@costello.watts2016]. According to the PT+N model, conditional probabilities $P(A|B)$ are estimated by a two-stage sampling procedure: first both events $A$ and $B$ are sampled with noise, and then a second noisy process computes the ratio of the events read as $A$ and $B$ over events read as $B$. The probability can be written as:

@@

Outside of conditional probability judgments, the primary difference between the models is the psychological process thought to produce bias in probability judgments: sample reading noise in the PT+N model and Bayesian inference in the Bayesian Sampler model. These process-level differences imply different response distributions.

In both models, the amount of variability in trial-level responses is related to the number of mental samples drawn, $N$. In the Bayesian Sampler model, assuming $\beta$ is relatively small, $N$ should also help to determine the degree to which responses are shrunk toward .50. In contrast, in the PT+N model the variance across responses and degree of shrinkage are reasonably considered to be independent. Moreover, the Bayesian Sampler model predicts not only shrinkage of probability estimates toward .50, but also a truncation of the response distribution in proportion to $\beta$ and $N$ [@chater.etal2020; @sundh.etal2021]. Thus, modeling the distributions of raw responses holds promise for disentangling the models.

However, there are at least three challenges to directly modeling raw human response data. First, both models are, strictly speaking, discrete and so make a limited set of discrete predictions while assigning zero probability to responses outside that set. Second, and similarly, the truncation in the Bayesian Sampler model also assigns zero probability to respones beyond the truncated range. And third, from a cursory glance it is clear that a majority of human responses are rounded by some unknown degree, with most seemingly rounded to the nearest 5 or 10%. Given the combination of these factors, if fit directly to raw human data the posterior probability of both models is likely to be zero. I return to these challenges and my approaches to addressing them in the results.

@@

Finally, Zhu and colleagues [-@zhu.etal2020] decision to fit "unpooled" estimates of all participants individually may exaggerate the complexity of both models. This is especially an issue in differentiating between simple and complex variants, which Zhu and colleauges [-@zhu.etal2020] elect to ignore by averaging the model scores. If there is limited heterogeneity across individuals, then adding a parameter for each participant may effectively over-penalize the complex variants relative to the simple variants. In addition, this could also affect comparisons across the PT+N and Bayesian Sampelr theories: if one model must assume greater heterogeneity in parameter estimates across people, then this model is more complex. Hierarchical models with partial pooling offer a solution that balances between ignoring individual variation and allowing it to vary freely, allowing for an accounting of heterogeneity without over-penalizing in cases where heterogeneity is low.

@@

Still, the results of this analysis were largely inconclusive with respect to differentiating the models [@zhu.etal2020]. Specifically, the quality of the fit for each model depended heavily on the maximum number of samples that is assumed possible. For small numbers of samples the Bayesian Sampler model is clearly superior, but for larger numbers of samples, the PT+N model was found to better fit the data. Presumably this is because with small numbers of samples the PT+N model is extremely constrained in the distinct discrete responses it can predict. For small N, both models predict only a limited set of distinct values are possible. However, whereas the size of that set is the same for each model, in the Bayesian Sampler the continuous $\beta$ parameter can shift exactly what those discrete values are, providing it much greater flexibility. However, this complexity goes unpunished in comparisons based on Wasserstein distance.

@@

Notice that for each model the probability judgments depend on underlying subjective probabilities, derived from a mental sampling process. These subjective probabilities are unobserved, and must be estimated as a latent variable. Here, they are represented with a four-dimensional dirichlet distribution for each subject, representing the probability of the elementary events ($A \land B$, $\neg A \land B$, $A \land \neg B$, $\neg A \land \neg B$).

@@

The function $f_{PT+N}$ computes the expected probability estimate using the underlying subjective probability $p$, the noise parameters $d$  and $d'$, and the relevant equation as defined by the PT+N theory (see supplemental materials for implementation details). Prior predictive checks were conducted for all models to select priors that would be uninformative or minimally informative on the scale of the model parameters $d$  and $d'$. [^3]

[^3]: Uninformativeness was sought in order to reduce bias in the posterior parameter estimates. It should be acknowledged that a uniform prior does not exactly correspond to what the authors of the PT+N theory would predict, as they have frequently assumed $d$ to be a fairly small value [e.g. @costello.watts2017]

Recall that Zhu and colleagues [-@zhu.etal2020] identified a bridging condition relating $\beta$ and $N$ in the Bayesian Sampler model to the $d$ parameter of the PT+N model. To support direct comparisons of the models, I parameterize the Bayesian Sampler model according to the implied $d$  and $d'$, rather than directly according to its $\beta$, $N$, and $N'$ parameters.^[Strictly speaking, under the original form of the Bayesian sampler model, $N$ and $N'$ are discrete parameters representing the number of distinct independent samples drawn. Given a particular implied $d$, this could create constraints on the possible values of $d'$, assuming $\beta$ is held constant. However, Zhu and colleagues [-@zhu.etal2020] also consider the possibility that people draw non-independent mental samples, in which case $N$ and $N'$ would represent the _effective number of samples_, accounting for their autocorrelation. In this case, we could treat this effective number of samples as a continuous quantity, and therefore imagine there are no clear constraints on $d$ and $d'$ except the stipulation that $d \leq d'$.] I constrain $d$ to $[0, 1/3]$ for the Bayesian Sampler model to reflect the assumption that $\beta \in [0, 1]$. This allows the same priors to be used for the corresponding Bayesian Sampler and PT+N models, simplifying their comparison.

@@

\begin{align*}
  \mu_{ijk} &= f_{BS}(\overrightarrow{\theta_{jk}}, d_j, d_j)  \\
  d_j &= \frac{1}{3} \ \text{logistic}(\delta_j) \\
  d_j’ &= \frac{1}{3} \ \text{logistic}\big(\delta_j + \exp(\Delta\delta_j)\big) \tag{10}
\end{align*}

@@

Recall, in contrast to a noise-based account, the Bayesian Sampler theory predicts both a truncation of the response distribution as well as a correlation between the degree of shrinkage in probability estimates and the variability of those estimates trial-to-trial (assuming $\beta$ is relatively small). Because the Bayesian Sampler applies a Bayesian adjustment after sampling, it predicts probability judgments will always lie between $d$ and $1 - d$, even when zero positive or negative mental samples are drawn (see Equation X and the bridging condition, Equation X). First, it bears noting that the truncated response distribution implied by the Bayesian Sampler model appears at odds with the raw response data. However, participants' responses frequently lie outside the range implied by the best estimates of their $d$ parameters (41% in Experiment 1 and 60% in Experiment 2).

@@

In the following set of analyses I attempt to lay out a set of reasonble assumptions that permit participants' trial-by-trial responses to be modeled and used to compare the theories' predictions. To do so, I first extend the models so as to render them fully continuous in their latent space. I then marry them with a specific model of response errors. Thus the models compared in the following analyses are not identical to those originally proposed by Zhu et al [-@zhu.etal2020] and Costello and Watts [-@costello.watts2014]. However, they do provide implementations of the theories' process-level accounts, and thus a means to test the distributional predictions of models based on Bayesian adjustment against models based on sampling noise.

@@

Zhu and collauges [-@zhu.etal2020] introduce the possibility of an "autocorrelated" Bayesian Sampler model under which samples are assumed to be autocorrelated (ideas which were advanced further in [@zhu.etal2021]). As autocorrelated samples provide less information than i.i.d samples, they should be weighted when computing probability estimates. The idea is that people actually draw N autocorrelated samples that approximate some smaller $N_{\text{eff}}$ i.i.d samples. Assuming the true number of samples drawn, N, is allowed to vary somewhat noisily, then a model based on autocorrelated samples would no longer be limited to predicting a discrete set of possible responses.

What distribution of mental samples would such a process form? Figure 8 shows the result of a simulation study examining probability estimates drawing on autocorrelated samples with estimated weights. Each simulation consists of a simple discrete Markov chain generating autocorrelated samples at $x_t$, returning $x_{t-1}$ with probability $p_{a}$, or else returning a new sample distributed Bernoulli($p$) was simulated. The setting of $p_a$ controls the degree of (positive) autocorrelation in the chain. Given the degree of autocorrelation in a sample, a weight is initially estimated as $N_{\text{eff}}/N$ and an appropriate number of samples $N$ are drawn to produce $N_{\text{eff}}$ samples. A probability estimate is then calculated from these samples. As the figure shows, for large-enough autocorrelation values ($p_a$) the resulting distribution becomes quite smooth, with the exception of spikes at zero and one for small to moderate values of $N_{\text{eff}}$. This occurs when all samples are zero or all one, because $\sum_i^N0 \cdot w_i = 0$ for any $w_i$ and $\sum_i^N 1 \cdot w_i = 1$ for any weights $\sum_i^N w_i = 1$.

@@

One possibility is that there is another source of error that explains these discrepancies, for instance that people just tend to avoid giving "extreme" responses. Another possibility is that the mental sampling process somehow enforces sampling of both positive and negative examples. A mental sampling process must start somewhere, and these initialization point(s) may influence judgments. For example, features of the initializations of mental sampling processes have been proposed to anchoring [@lieder.etal2013] and unpacking effects [@dasgupta.etal2017; also see @sanborn.chater2016]. In this spirit, one possible explanation for the relative absence of extreme responses is that people draw samples following two initializations, initializing sampling with both a positive and negative instance. As the simulations in Figure 8 (bottom) show, this produces a predicted response distribution that is  approximately $Beta(1 + p N_{\text{eff}}, 1+(1-p)N_{\text{eff}})$.

Incorporating these assumptions provides a psychologically plausible mechanism to permit the use of a Beta distribution to approximate the latent sampling process, rendering both of the models' latent parameters continuous. Note that this process will also itself produce some  shrinkage in the probability estimates, and this shrinkage will be somewhat sensitive to the number of samples drawn in both the Bayesian Sampler and PT+N models. Nevertheless the dependency between shrinkage and variance remains stronger for the Bayesian Sampler model and it is only the Bayesian Sampler model that predicts a truncation of the response distribution.

@@

The Bayesian Sampler model predicts a truncated range of possible responses given $\beta$ and $N$ (and consequently, implied $d$). Modeling these different rounding processes allows for at least some out-of-bounds responses to be accounted for by rounding processes (e.g. when an allowable response of .14 is rounded to the out-of-bounds value of .10). However, some participant response still cannot be accounted for by the model. Instead, these responses are treated as "contaminants" generated by a random response process. Modeling "contaminant" response processes allows the Bayesian Sampler model to be fit in the presence of true outliers. Identifying the estimated proportion of "contaminant" responses can also provide a check on the models: if a model can only be fit by assuming a large proportion of contaminant responses, this suggests it is likely not a good model of human behavior.

@@

Compared to the query-averaged model, the trial-level noise-based model adds two features: mixture components for rounding and contaminants and subject-level varying $N$ in place of a fixed $K$ parameter. This model's implementation and the implementation of its mixture components is depicted in Figure 9. However, note that $N$ is allowed to vary independently from $d$, allowing for independence between response shrinkage and variability. Moreover, the latent distribution used to calculate categorical cut points is an untransformed Beta distribution.

@@

Table 2 shows the scores of each model fit to the trial-level data in Experiments 1 and 2. From the quantitative model comparison it is clear that the noise-based model is superior, with substantialy lower $\widehat{\text{elpd}}_{\text{LOO}}$ than both of the competing Bayesian Sampler implementations in both experiments.>

@@

Second, we can examine the shrinkage-variance relationship. In the noise-based model, N and d were allowed to vary freely. But if these quantities indeed are actually correlated as predicted by a Bayesian adjustment model, then we should expect to nevertheless see a correlation between the subject-level estimates in these parameters. Figure 11 shows a scatterplot relating these estimates. Looking across subjects we see they are mostly uncorrelated. Although there is a slight negative correlation in Experiment 1, this is driven by only a handful of participants with extreme values and in Experiment 2 the pattern appears to if anything be reversed. These findings again run counter to the predictions of the Bayesian Sampler theory.

@@

Comparing the d values inferred from the trial-level model we see they are similar to but generally smaller than those estimated in the trial-averaged model. This is likely because the hypothesized initialization process accounts for some of the shrinkage in participants' responses, so that less need be attributed to sample read noise.

@@

Fit to the average of participants' responses over blocks, there is a single clear winner: a model without any special treatment of conditional probability (ala the Bayesian Sampler model) and allowing for an implied $d$ parameter $\in [0,.5]$. This model could be interpreted either as a variant of the Bayesian Sampler without restriction on its $\beta$ parameter, or as a variant of the PT+N model that removes its account of conditional probability judgments.

@@

Directly fitting the models to these trial-level human response data presents several challenges addressed here by creating continuous extensions of these models and incorporating additional modeling of response processes. Though not identical to the models as originally proposed, this exercise allowed for tests of the key distributional predictions of the theories. Neither of the Bayesian Sampler's predictions appear to be borne out by the data. First, participants' responses frequently fall outside the truncated range implied by the parameters estimated under the Bayesian Sampler model. Fit to the raw data, this requires treating an unreasonably large proportion of responses as "contaminants". Second, the degree of shrinkage in participants' responses and the variability in those responses appear largely uncorrelated.

@@

High estimates of $d$ parameters might also call into question arguments for the rational utility of such a process. Zhu and colleagues argue that the regularizing effect of bayesian adjustment should be seen as adaptive. They also consider that "noise" might give an algorithmic-level solution to the computation-level goals defined by the Bayesian Sampler [-@zhu.etal2020]. Even high implied $d$ values might be consistent with rational inference in cases where the number of effective mental samples is very low. For instance, a beta(2,2) prior is only modestly informative, but could produce $d = .40$ if $N=1$. However, if more samples are drawn then high $d$ values would correspond to potentially inflexible and suboptimal priors (i.e. too large $\beta$ values). From the results, it is clear there are some individuals with both relatively high $d$ and $N$ values, which may press somewhat against the rational justification for the desirability of sampling noise.

@@

Finally, one important direction for future elaborations of sampling-based theories are more rigorous theories of realistic mental sampling processes, including details of the initialization and autocorrelation among samples. Here, I proposed a relatively-highly autocorrelated sampling process initialized with a positive and negative example to account for qualitative features of participants' response distributions and to support a simple approximation. Future work might explore the details of these processes and incorporate them into models of probability judgments in more rigorous ways.

>>>>

keywords          : "probability judgments, Bayesian cognitive science, heuristics and biases"
wordcount         : "9195"

@@

By assumption, a maximum of 50% of samples can be misread on average, so $d$ is a number in the range $[0, 1/2]$. The overall consequence of the sample-reading noise will be to shrink probability estimates toward .50 in proportion to $d$. The PT+N theory provides a unified account for a wide variety of biases in probability judgment that were previously attributed to different types of heuristics, as well as novel biases identified based on the model's predictions [@costello.watts2014; @costello.watts2016; @costello.watts2017; @costello.watts2018]. For example, the PT+N theory offers an explanation for many instances of "conservatism" [@costello.watts2014]---people's tendency to shy away from extreme probability judgments near 0 and 1, even when strong evidence warrants such judgments [e.g. @edwards1968; @erev.etal1994].

@@

The model's predictions can be distinguished on two levels: First, the models have distinct accounts of conditional probability judgments that make different predictions in terms of expected values. Second, the models present different process-level accounts of probability judgment that entail different predictions about the shape of the distribution of responses across individuals.

@@

In contrast, the PT+N model presents a constructive account of conditional probability judgments that is fundamentally non-Bayesian [@costello.watts2016]. According to the PT+N model, conditional probabilities $P(A|B)$ are estimated by a two-stage sampling procedure: first both events $A$ and $B$ are sampled with noise, and then a second noisy process computes the ratio of the events read as $A$ and $B$ over events read as $B$. Schematically, the estimated probability can be written as:

@@

Although the model's predictions for unconditional probability judgments are identical in expectation (as seen via the bridging condition), the models posit different psychological processes underlying those judgments: sample reading noise in the PT+N model and Bayesian inference in the Bayesian Sampler model. These process-level differences imply different predictions about the distributions of people's judgments.

The models make qualitatively different distributional predictions on two fronts. First, the Bayesian Sampler predicts a clear relationship between the degree to which responses are shrunk toward .50 and the trial-by-trial variability in those responses. In both models, the amount of variability in trial-level responses is related to the number of mental samples drawn, $N$. In the Bayesian Sampler model, assuming $\beta$ is relatively small, $N$ should also help to determine the degree to which responses are shrunk toward .50. In contrast, in the PT+N model the variance across responses and degree of shrinkage are reasonably considered to be independent. Second, because the Bayesian Sampler describes a process of adjustment after the sampling process, in which people report the mean of their mental posterior over probabilities, the model also predicts a truncation of the response distribution in proportion to $\beta$ and $N$ [@chater.etal2020; @sundh.etal2021]. That is, even when zero positive or negative samples are drawn, the mean of the posterior is drawn away from extreme responses of zero and one.

Modeling the distributions of raw responses holds clear promise for disentangling the models. However, there are at least three challenges to directly modeling raw human response data. First, both models are, strictly speaking, discrete and so make a limited set of discrete predictions while assigning zero probability to responses outside that set. Second, and similarly, the truncation in the Bayesian Sampler model also assigns zero probability to respones beyond the truncated range. And third, from a cursory glance it is clear that a majority of human responses are rounded by some unknown degree, with most seemingly rounded to the nearest 5 or 10%. Given the combination of these factors, if fit directly to raw human data the posterior probability of both models is likely to be zero. I return to these challenges and my approaches to addressing them in the results.

@@

Finally, Zhu and colleagues [-@zhu.etal2020] decision to fit "unpooled" estimates of all participants individually may exaggerate the complexity of both models. This is especially likely to pose an issue in differentiating between simple and complex variants, which Zhu and colleauges [-@zhu.etal2020] elect to ignore by averaging the model scores. If there is limited heterogeneity across individuals, then adding a parameter for each participant may effectively over-penalize the complex variants relative to the simple variants. In addition, this could also affect comparisons across the PT+N and Bayesian Sampelr theories: if one model must assume greater heterogeneity in parameter estimates across people, then this model is more complex. Hierarchical models with partial pooling offer a solution that balances between ignoring individual variation and allowing it to vary freely, allowing for an accounting of heterogeneity without over-penalizing in cases where heterogeneity is low.

@@

Still, the results of this analysis were largely inconclusive with respect to differentiating the models [@zhu.etal2020]. Specifically, the quality of the fit for each model depended heavily on the maximum number of samples that is assumed possible. For small numbers of samples the Bayesian Sampler model is clearly superior, but for larger numbers of samples, the PT+N model was found to better fit the data. Presumably this is because with small numbers of samples the PT+N model is extremely constrained in the distinct discrete responses it can predict. For small N, both models predict only a limited set of distinct values are possible. However, whereas the size of that set is the same for each model, in the Bayesian Sampler the continuous $\beta$ parameter can shift exactly what those discrete values are, providing it much greater flexibility. Yet, this additional model flexibility goes unpunished in comparisons based on Wasserstein distance.

@@

Notice that for each model the probability judgments depend on underlying subjective probabilities, derived from a mental sampling process. These subjective probabilities are unobserved, and must be estimated as a latent variable. Here, they are represented with a four-dimensional dirichlet distribution for each subject, $\vec{\theta}$, representing the probability of the elementary events ($A \land B$, $\neg A \land B$, $A \land \neg B$, $\neg A \land \neg B$).

@@

The function $f_{PT+N}$ computes the expected probability estimate using the underlying subjective probability computed from $\vec{\theta}$ and the query, the noise parameters $d$  and $d'$, and the relevant equation as defined by the PT+N theory (see supplemental materials for implementation details). Prior predictive checks were conducted for all models to select priors that would be uninformative or minimally informative on the scale of the model parameters $d$  and $d'$. [^3]

[^3]:Uninformativeness was sought in order to reduce bias in the posterior parameter estimates. It should be acknowledged that a uniform prior does not exactly correspond to what the authors of the PT+N theory would predict, as they have frequently assumed $d$ to be a fairly small value [e.g. @costello.watts2017]

Recall that Zhu and colleagues [-@zhu.etal2020] identified a bridging condition relating $\beta$ and $N$ in the Bayesian Sampler model to the $d$ parameter of the PT+N model. To support direct comparisons of the models, I parameterize the Bayesian Sampler model according to the implied $d$  and $d'$, rather than directly according to its $\beta$, $N$, and $N'$ parameters.[^4] I constrain $d$ to $[0, 1/3]$ for the Bayesian Sampler model to reflect the assumption that $\beta \in [0, 1]$. This allows the same priors to be used for the corresponding Bayesian Sampler and PT+N models, simplifying their comparison.

[^4]:Strictly speaking, under the original form of the Bayesian sampler model, $N$ and $N'$ are discrete parameters representing the number of distinct independent samples drawn. Given a particular implied $d$, this could create constraints on the possible values of $d'$, assuming $\beta$ is held constant. However, Zhu and colleagues [-@zhu.etal2020] also consider the possibility that people draw non-independent mental samples, in which case $N$ and $N'$ would represent the _effective number of samples_, accounting for their autocorrelation. In this case, we could treat this effective number of samples as a continuous quantity, and therefore imagine there are no clear constraints on $d$ and $d'$ except the stipulation that $d \leq d'$. These ideas will be developed further in the trial-level analyses.

@@

\begin{align*}
  \mu_{ijk} &= f_{BS}(\overrightarrow{\theta_{jk}}, x_{ijk}, d_j, d'_j)  \\
  d_j &= \frac{1}{3} \ \text{logistic}(\delta_j) \\
  d_j’ &= \frac{1}{3} \ \text{logistic}\big(\delta_j + \exp(\Delta\delta_j)\big) \tag{10}
\end{align*}

@@

Recall, in contrast to a noise-based account, the Bayesian Sampler theory predicts both a truncation of the response distribution as well as a correlation between the degree of shrinkage in probability estimates and the variability of those estimates trial-to-trial (assuming $\beta$ is relatively small). Because the Bayesian Sampler applies a Bayesian adjustment after sampling, it predicts probability judgments will always lie between $d$ and $1 - d$, even when zero positive or negative mental samples are drawn (see Equation 12 and the bridging condition, Equation 5). First, it bears noting that the truncated response distribution implied by the Bayesian Sampler model appears at odds with the raw response data. However, participants' responses frequently lie outside the range implied by the best estimates of their $d$ parameters (41% in Experiment 1 and 60% in Experiment 2).

@@

In the following set of analyses I attempt to lay out a set of reasonble assumptions that permit participants' trial-by-trial responses to be modeled and used to compare the theories' predictions. To do so, I first extend the models so as to render them fully continuous in their latent space and then marry them with a specific model of response errors. Thus the models compared in the following analyses are not identical to those originally proposed by Zhu et al [-@zhu.etal2020] and Costello and Watts [-@costello.watts2014]. However, they do provide implementations of the theories' process-level accounts, and thus a means to test the distributional predictions of models based on Bayesian adjustment against models based on sampling noise.

@@

Zhu and collauges [-@zhu.etal2020] introduce the possibility of an "autocorrelated" Bayesian Sampler model under which samples are assumed to be autocorrelated [ideas which were advanced further in [@zhu.etal2021]. As autocorrelated samples provide less information than i.i.d samples, they should be weighted when computing probability estimates. The idea is that people actually draw N autocorrelated samples that approximate some smaller $N_{\text{eff}}$ i.i.d samples. Assuming the true number of samples drawn, N, is allowed to vary somewhat noisily, then a model based on autocorrelated samples would no longer be limited to predicting a discrete set of possible responses.

What distribution of mental samples would such a process form? Figure 8 shows the result of a simulation study examining probability estimates drawing on autocorrelated samples with estimated weights. Each simulation consists of a simple discrete Markov chain generating autocorrelated samples at $x_t$, returning $x_{t-1}$ with probability $p_{a}$, or else returning a new sample distributed Bernoulli($p$) was simulated. The setting of $p_a$ controls the degree of (positive) autocorrelation in the chain. Given the degree of autocorrelation in a sample, a weight is initially estimated as $N_{\text{eff}}/N$ and an appropriate number of samples $N$ are drawn to produce $N_{\text{eff}}$ samples. A probability estimate is then calculated from these samples. As the figure shows, for large-enough autocorrelation values ($p_a$) the resulting distribution becomes quite smooth, with the exception of spikes at zero and one for small to moderate values of $N_{\text{eff}}$. This occurs when all samples are zero or all one.

@@

One possibility is that there is another source of error that explains these discrepancies, for instance that people just tend to avoid giving "extreme" responses. Another possibility is that the mental sampling process somehow enforces sampling of both positive and negative examples. A mental sampling process must start somewhere, and these initialization point(s) may influence judgments. For example, features of the initializations of mental sampling processes have been proposed to anchoring [@lieder.etal2012] and unpacking effects [@dasgupta.etal2017; also see @sanborn.chater2016]. In this spirit, one possible explanation for the relative absence of extreme responses is that people draw samples following two initializations, initializing sampling with both a positive and negative instance. As the simulations in Figure 8 (bottom) show, this produces a predicted response distribution that is  approximately $Beta(1 + p N_{\text{eff}}, 1+(1-p)N_{\text{eff}})$.

Assuming such an autocorrelated and initialized sampling process provides a psychologically plausible mechanism to permit the use of a Beta distribution to approximate the latent sampling process, rendering both of the models' latent parameters continuous. Note that this process will also itself produce some shrinkage in the probability estimates, and this shrinkage will be somewhat sensitive to the number of samples drawn in both the Bayesian Sampler and PT+N models. Nevertheless the dependency between shrinkage and variance remains stronger for the Bayesian Sampler model and it is only the Bayesian Sampler model that predicts a truncation of the response distribution.

@@

The Bayesian Sampler model predicts a truncated range of possible responses given $\beta$ and $N$ (and consequently, implied $d$). Modeling these different rounding processes allows for at least some out-of-bounds responses to be accounted for by rounding processes (e.g. when an allowable response of .14 is rounded to the out-of-bounds value of .10). However, some responses still cannot be accounted for by the model. Instead, these responses are treated as "contaminants" generated by a random response process. Modeling "contaminant" response processes allows the Bayesian Sampler model to be fit in the presence of true outliers. Identifying the estimated proportion of "contaminant" responses can also provide a check on the models: if a model can only be fit by assuming a large proportion of contaminant responses, this suggests it is likely not a good model of human behavior.

@@

Compared to the query-averaged model, the trial-level noise-based model adds two features: mixture components for rounding and contaminants and subject-level varying $N$ in place of a fixed $K$ parameter. This model's implementation and the implementation of its mixture components is depicted in Figure 9. However, note that $N$ is allowed to vary independently from $d$, allowing for independence between response shrinkage and variability.

@@

Table 2 shows the scores of each model fit to the trial-level data in Experiments 1 and 2. From the quantitative model comparison it is clear that the noise-based model is superior, with substantialy lower $\widehat{\text{elpd}}_{\text{LOO}}$ than both of the competing Bayesian Sampler implementations in both experiments.

@@

Second, we can examine the shrinkage-variance relationship. In the noise-based model, $N$ and $d$ were allowed to vary freely. But if these quantities indeed are actually correlated as predicted by a Bayesian adjustment model, then we should expect to nevertheless see a correlation between the subject-level estimates in these parameters. Figure 11 shows a scatterplot relating these estimates. Looking across subjects we see they are mostly uncorrelated. Although there is a slight negative correlation in Experiment 1, this is driven by only a handful of participants with extreme values and in Experiment 2 the pattern appears to if anything be reversed. These findings again run counter to the predictions of the Bayesian Sampler theory.

@@

Comparing the $d$ values inferred from the trial-level model we see they are similar to but generally smaller than those estimated in the trial-averaged model. This is likely because the hypothesized initialization process accounts for some of the shrinkage in participants' responses, so that less need be attributed to sample read noise.

@@

Fit to the average of participants' responses over blocks, there is a single clear winner among the competing models: a model without any special treatment of conditional probability (ala the Bayesian Sampler model) and allowing for an implied $d$ parameter $\in [0,.5]$. This model could be interpreted either as a variant of the Bayesian Sampler without restriction on its $\beta$ parameter, or as a variant of the PT+N model that removes its account of conditional probability judgments.

@@

Directly fitting the models to these trial-level human response data presents several challenges addressed here by creating continuous extensions of these models and incorporating additional modeling of response processes. Though not identical to the models as originally proposed, this exercise allowed for tests of the key distributional predictions of the theories. Neither of the Bayesian Sampler's predictions appear to be borne out by the data. First, participants' responses frequently fall outside the truncated range implied by the parameters estimated under the Bayesian Sampler model. Fit to the raw data, this requires treating an unreasonably large proportion of responses as "contaminants". Second, the degree of shrinkage in participants' responses and the variability in those responses are not correlated in the ways predicted by the Bayesian Sampler theory.

@@

High estimates of $d$ parameters might also call into question arguments for the rational utility of such a process. Zhu and colleagues argue that the regularizing effect of bayesian adjustment should be seen as adaptive. They also consider that "noise" might give an algorithmic-level solution to the computation-level goals defined by the Bayesian Sampler [-@zhu.etal2020]. Even high implied $d$ values might be consistent with rational inference in cases where the number of effective mental samples is very low. For instance, a Beta(2,2) prior is only modestly informative, but could produce $d = .40$ if $N=1$. However, if more samples are drawn then high $d$ values would correspond to potentially inflexible and suboptimal priors (i.e. too large $\beta$ values). From the results, it is clear there are some individuals with both relatively high $d$ and $N$ values, which may press somewhat against the rational justification for the desirability of sampling noise.

@@

Finally, one important direction for future elaborations of sampling-based theories are more rigorous theories of realistic mental sampling processes, including details of their initialization, autocorrelation, and amortization [@gershman.goodman2016]. Here, I proposed an autocorrelated sampling process initialized with a positive and negative example to account for qualitative features of participants' response distributions and to support a simple approximation. Future work might explore the details of these processes and incorporate them into models of probability judgments in more rigorous ways.
